---
title: "The Impact of Height on NHL Goalies (Data Extraction)"
author: "Rayce Ramsay"
date: "2025-03-13 (Updated 2025-04-27)"
output: github_document
---


## Resources

- [(Unofficial) NHL API Docs](https://github.com/Zmalski/NHL-API-Reference?tab=readme-ov-file)
- [NHL Goalie Bios API Example](https://api.nhle.com/stats/rest/en/goalie/bios?limit=-1&start=&sort=&cayenneExp=seasonId=20232024)
- [Official NHL Goalie Stats](https://www.nhl.com/stats/goalies?report=bios&reportType=season&seasonFrom=20242025&seasonTo=20242025&gameType=2&sort=a_goalieFullName&page=0&pageSize=100)


## Setup

```{r, message=FALSE, warning=FALSE}
# Load libraries
library(tidyverse)
library(httr2)
```

```{r}
# Base URLS for NHL API
NHL_API_BASE_URL = "https://api.nhle.com/stats/rest/"
NHLWEB_API_BASE_URL = "https://api-web.nhle.com/"
```

```{r}
# Get season IDs to query NHL API for starting with the 2000-2001 season
START_YEARS = 2000:2024
season_ids = NULL
for (start_year in START_YEARS) {
  season_ids = c(season_ids, paste0(start_year, start_year + 1))
}

season_ids
```


## Get NHL Goalie Season Data

```{r}
# Create URLs to fetch NHL goalie info by season
summary_urls = paste0(NHL_API_BASE_URL, "en/goalie/summary?limit=-1&start=&sort=&cayenneExp=seasonId=", season_ids)
bio_urls = paste0(NHL_API_BASE_URL, "en/goalie/bios?limit=-1&start=&sort=&cayenneExp=seasonId=", season_ids)

# Create http request objects
bio_reqs = lapply(bio_urls, httr2::request)
summary_reqs = lapply(summary_urls, httr2::request)

# Perform http requests
bio_resps = req_perform_sequential(bio_reqs)
summary_resps = req_perform_sequential(summary_reqs)
```

```{r}
# Extract bios and summaries from requests
bio_data = map2(bio_resps, season_ids, function(resp, season) {
  goalies = resp_body_json(resp)$data
  lapply(goalies, function(goalie) {
    goalie$seasonId = season
    map(goalie, ~ifelse(is.null(.), NA, .))
  })
})
summary_data = map(summary_resps, function(resp) {
  goalies = resp_body_json(resp)$data
  lapply(goalies, function(goalie) {
    goalie$seasonId = as.character(goalie$seasonId)
    map(goalie, ~ifelse(is.null(.), NA, .))
  })
})

# Convert extracted data to tibbles
bios = list_rbind(map(list_c(bio_data), ~as_tibble(.x)))
summaries = list_rbind(map(list_c(summary_data), ~as_tibble(.x)))

dim(bios)
head(bios)
tail(bios)
dim(summaries)
head(summaries)
tail(summaries)
```

```{r}
# Select columns from summaries to keep for joining
summaries_limited = summaries |>
  select(!c(goalieFullName, lastName, shootsCatches)) # only keep summary/stats (non-bio) columns

# Join goalie biographies and summaries into one tibble
goalies = bios |>
  select(!c(currentTeamAbbrev, firstSeasonForGameType, gamesPlayed, losses, otLosses, shutouts, ties, wins)) |> # only keep bio (non-summary) columns
  distinct() |> # ignore duplicates since data has separate rows for playoffs and regular season
  inner_join(summaries_limited, by = join_by(seasonId, playerId))
  
dim(goalies)
head(goalies)
tail(goalies)
```

```{r}
# Tidy up some columns and sort
goalies = goalies |>
  rename(
    catches = shootsCatches,
    fullName = goalieFullName
  ) |>
  arrange(seasonId, fullName)

dim(goalies)
head(goalies)
tail(goalies)
```

```{r}
# Export NHL goalies data to a csv
write_csv(goalies, "../data/nhl_goalies_data.csv")
```


## Get NHL Draft Ranking Data

```{r}
DRAFT_YEARS = 2008:2025  # data only goes back until 2008

# Create URLs to fetch goalie prospect rankings by draft year.
# Notice there are separate urls for North American and International goalies
na_urls = paste0(NHLWEB_API_BASE_URL, "v1/draft/rankings/", DRAFT_YEARS, "/3")
itl_urls = paste0(NHLWEB_API_BASE_URL, "v1/draft/rankings/", DRAFT_YEARS, "/4")

# Create http request objects
na_reqs = lapply(na_urls, httr2::request)
itl_reqs = lapply(itl_urls, httr2::request)

# Perform http requests
na_resps = req_perform_sequential(na_reqs)
itl_resps = req_perform_sequential(itl_reqs)
```

```{r}
# Extract north american and international goalie prospects from requests
na_data = map(na_resps, function(resp) {
  resp_data = resp_body_json(resp)
  goalies = resp_data$rankings
  lapply(goalies, function(goalie) {
    goalie$draftYear = resp_data$draftYear
    goalie$midtermRank = ifelse(goalie$midtermRank, goalie$midtermRank, NA_integer_)
    goalie$finalRank = ifelse(goalie$finalRank, goalie$finalRank, NA_integer_)
    goalie$international = FALSE
    map(goalie, ~ifelse(is.null(.), NA, .))
  })
})
itl_data = map(itl_resps, function(resp) {
  resp_data = resp_body_json(resp)
  goalies = resp_data$rankings
  lapply(goalies, function(goalie) {
    goalie$draftYear = resp_data$draftYear
    goalie$midtermRank = ifelse(goalie$midtermRank, goalie$midtermRank, NA_integer_)
    goalie$finalRank = ifelse(goalie$finalRank, goalie$finalRank, NA_integer_)
    goalie$international = TRUE
    map(goalie, ~ifelse(is.null(.), NA, .))
  })
})

# Convert extracted data to tibbles
na_prospects = list_rbind(map(list_c(na_data), \(x) as_tibble(x) ))
itl_prospects = list_rbind(map(list_c(itl_data), \(x) as_tibble(x) ))

dim(na_prospects)
head(na_prospects)
tail(na_prospects)
dim(itl_prospects)
head(itl_prospects)
tail(itl_prospects)
```

```{r}
# Join north american and international goalie prospects into one tibble
prospects = rbind(na_prospects, itl_prospects)

dim(prospects)
```

```{r}
# Tidy up some columns and sort
prospects = prospects |>
  mutate(
    fullName = paste(firstName, lastName)
  ) |>
  select(!c(firstName, lastName)) |>
  rename(
    catches = shootsCatches,
    height = heightInInches,
    weight = weightInPounds, 
    birthCountryCode = birthCountry
  ) |>
  arrange(draftYear, finalRank, midtermRank, fullName)

dim(prospects)
head(prospects)
tail(prospects)
```

```{r}
# Export NHL prospects data to a csv
write_csv(prospects, "../data/nhl_goalie_prospects_data.csv")
```


*The following code was added on 2025-04-27.*

## Get NHL Season Standings Data

We need to first query an endpoint to find the last day of the regular season from each season so we can figure out end of season standings for each team later.

```{r}
# Create URL to fetch NHL structure data for each season 
struct_url = paste0(NHLWEB_API_BASE_URL, "v1/standings-season")

# Create http request object
struct_req = httr2::request(struct_url)

# Perform http request
struct_resp = req_perform(struct_req)
```

```{r}
# Extract season structures from request
struct_data = resp_body_json(struct_resp)$seasons

# Convert extracted data to tibble
structs = list_rbind(map(struct_data, ~as_tibble(.x)))

# Only keep season ids and end dates past 2000
structs = structs |>
  rename(seasonId = id) |>
  filter(seasonId %in% season_ids) |>
  select(seasonId, standingsEnd) |>
  arrange(seasonId)

# Change 2024-25 season standings date to match previous extraction date (if running later than previous code)
structs = structs |>
  mutate(standingsEnd = if_else(seasonId == 20242025, "2025-03-13", standingsEnd))

dim(structs)
structs
```

Now that we know the last day of each, we can query for final team standings.

```{r}
# Create URLs to fetch NHL team standings info by season
standings_urls = paste0(NHLWEB_API_BASE_URL, "v1/standings/", structs$standingsEnd)

# Create http request objects
standings_reqs = lapply(standings_urls, httr2::request)

# Perform http requests
standings_resps = req_perform_sequential(standings_reqs)
```

```{r}
# Extract team standings per season from requests
standings_data = map(standings_resps, function(resp) {
  standings = resp_body_json(resp)$standings
  lapply(standings, function(standing) {
    standing$placeName = standing$placeName$default
    standing$teamName = standing$teamName$default
    standing$teamCommonName = standing$teamCommonName$default
    standing$teamAbbrev = standing$teamAbbrev$default
    map(standing, ~ifelse(is.null(.), NA, .))
  })
})

# Convert extracted data to tibbles
standings = list_rbind(map(list_c(standings_data), ~as_tibble(.x)))

# Keep relevant team info and stats
standings = standings |>
  select(seasonId, teamAbbrev, pointPctg, winPctg, goalsForPctg, goalDifferentialPctg, wildcardSequence) |>
  arrange(seasonId, teamAbbrev)

dim(standings)
head(standings)
tail(standings)
```

```{r}
# Export NHL standings data to a csv
write_csv(standings, "../data/nhl_standings_data.csv")
```


## Get NHL Team Season Data

The following team datasets were downloaded manually for each season from https://moneypuck.com/data.htm.

```{r}
# Load in team stats datasets from each year (only have data since 2008-09 season)
TEAM_START_YEARS = 2008:2024
team_data_paths = paste0("../data/raw/teams_", TEAM_START_YEARS, ".csv")
teams = team_data_paths |>
  map(read_csv) |>
  bind_rows()

head(teams)
tail(teams)
```


```{r}
# View column names for teams data
dim(teams)
names(teams)
```


```{r}
# Tidy up metadata columns and only keep relevant info
teams = teams |>
  filter(situation == "all", position == "Team Level") |>
  mutate(
    seasonId = as.integer(paste0(as.character(season), season + 1)),
    teamAbbrev = as.factor(team...1)
  ) |>
  select(seasonId, teamAbbrev, iceTime, xGoalsAgainst, blockedShotAttemptsAgainst, dZoneGiveawaysAgainst)
  
dim(teams)
head(teams)
tail(teams)
```

```{r}
# View differences in team name abbreviations between datasets
table(sort(as.character(teams$teamAbbrev)))
table(sort(as.character(goalies$teamAbbrevs)))
```

```{r}
# Rename team name abbreviations from teams to match goalies dataset
teams = teams |>
  mutate(
    teamAbbrev = case_when(
      teamAbbrev == "L.A" ~ "LAK",
      teamAbbrev == "N.J" ~ "NJD",
      teamAbbrev == "S.J" ~ "SJS",
      teamAbbrev == "T.B" ~ "TBL",
      TRUE ~ teamAbbrev
    )
  )

table(teams$teamAbbrev)
```

```{r}
# Export NHL teams data to a csv
write_csv(teams, "../data/nhl_teams_data.csv")
```
